Inequality in socioemotional skills: a cross-cohort comparison

### 2020-02-11 codes/R/paper/text/paper_jpube_2datamethods.tex
Gabriella's text updates

We use information from two nationally representative longitudinal studies in the UK. The studies follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) surveys individuals born in 1970, and the Millennium Cohort Study (MCS) includes births between 2000 and 2002. The British Cohort Study includes all individuals born across Great Britain in a single week in 1970. Cohort members' families -- and subsequently members themselves -- were surveyed on multiple occasions. For this paper we augment the information at the five-year survey with data from birth, adolescence (16), and adulthood (30, 38, 42). The Millennium Cohort Study follows individuals born in the UK between September 2000 and January 2002. We use the first (age 0) survey -- carried out at 9 months of age -- and the sweeps at around 5 and 14 years of age.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}

@@

Ideally, we would compare socio-emotional skills alongside cognitive skills. However, the cognitive tests administered to each cohort have no overlap, even at the item level. We thus use the available cognitive tests in each cohort to estimate simple confirmatory factor model with a single latent dimension, separately by cohort (see \autoref{tab:harmvar} for the tests used). Unlike the other indicators in our analysis, cognitive skills are thus not comparable across cohort.

Another complication arises from the fact that, differently from the British Cohort Study, the Millennium Cohort Study has a stratified design. It oversamples children living in administrative areas characterised by higher socioeconomic deprivation and larger ethnic minority population \citep{Plewis2007}. We rebalance the MCS sample to make it nationally representative by excluding from the analysis a fraction of observations from the oversampled areas, proportionally to their sampling probability.\footnote{See Table 5.5 in \citet{Plewis2007}. This choice is mainly driven by software limitations. The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.} Finally, we also restrict our sample to individuals born in England and to cases where there is complete information on socioemotional skills at five years of age. The final sample contains 9,545 individuals from the British Cohort Study, and 5,572 from the Millennium Cohort Study. Summary statistics for the full and estimation samples are displayed in \autoref{tab:sumstats}. After the rebalancing step, the MCS estimation sample closely mirrors the full sample in terms of average observable characteristics, thus preserving representativeness.

@@

Child socio-emotional skills are an unobservable and difficult to measure construct. Over recent years, the measurement of such skills has evolved and, over time, different measures have been used.  As we discuss below, this makes the comparison  of socio emotional skills across different groups, assessed with different tools, difficult.

A common approach to infer a child's socio-emotional development is based on behavioural screening scales. As part of these tools, mothers (or teachers) indicate whether their children exhibit a series of behaviours -- the \emph{items} of the scale. In the British and Millennium Cohort Studies, two different scales were employed. In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS cohort, mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an update to the Rutter scale. It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}. \citet{Goodman1997} administered both scales to a sample of children, and showed that the scores are highly correlated, and the two measures do not differ in their discriminatory ability. The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively. In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health. However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{Goodman2011}.

While the Rutter and SDQ scales are similar in their components (since the latter was developed from the former, see \citet{Goodman1994}), there is no a priori reason to expect them to be directly comparable. First, the overlap of behaviours described in the two scales is only partial, given that - by design - the SDQ includes also strengths, in addition to weaknesses. Second, the wording of each item is slightly different, both in the description and in the options that can be selected as answers. Third, the different ordering of the items within each scale might lead to order effects. Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2006) might differ due to a host of evolving societal norms. Nonetheless, the level of comparibility of the two scales is higher than that of other scales used in comparating work in the literature reviewed in section \ref{sec:lit}.

As our goal is to compare socio-emotional skills across the two cohorts, we construct a new scale by retaining the items that are worded in a similar way across the two original Rutter and SDQ scales, and making some slight coding adjustments to maximise comparability. In what follows, we will consider the included items to be the same \emph{measure} in the two cohorts. The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability. We exclude from the analysis items that were completely different between the two questionnaires to maximise comparability between the two cohorts, as it is standard good practice in the psychometric literature (see for example \citet{Kern2014}).\footnote{Of course, we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.} More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.

Item-level prevalence by cohort and gender is in \autoref{tab:meantab}. We see that, in general, item prevalence is more similar across genders within the same cohort, than across cohorts. For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2006 than in 1975. Regardless, a simple cross-cohort comparison of item-level prevalence is misleading because of changing perceptions and norms about what constitutes problematic behaviour in children. The analysis in section \ref{sec:measinv} tackles this issue.

In the remainder of this section, we analyse the properties of the new scale. Following a common approach, we proceed in two steps. First, we carry out an \emph{exploratory} step, where we study the factor structure of our scale. The aim of this step is to examine the correlation between observed measures in a data-driven way, imposing the least possible assumptions. Here, we establish how many latent dimensions of socio-emotional skills the scale is capturing, and which items of our scale are measuring which dimension. As a second step, we set up a \emph{confirmatory} factor model. This model fixes the number of latent dimensions, and imposes a dedicated measurement structure, based on the insights obtained in the exploratory step. This is the model from which the measurement invariance analysis of \autoref{sec:measinv} departs.

@@

The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}. This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \citet{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}. However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each. The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}. This five-factor structure has been validated in many contexts \citep{Stone2010}; lower-dimensional structures have been also suggested \citep{Dickey2004}. Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}. Indeed, the internalising and externalising dimensions were introduced in psychology by \citet{Achenbach1966}, who show that they are the two main factors underlying a wide range of psychological measures; as noted in \citet{Achenbach2016}, more than 75,000 articles have been published on internalizing and externalizing problems.

@@

Measurement invariance analysis necessarily starts from a minimally restrictive model, denominated \emph{configural} model. This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts. It serves as a basis for our measurement invariance analysis.

The restrictions implied in \eqref{eq:propens} and \eqref{eq:thresh} are not sufficient to identify the parameters of the model. Even with these assumptions, there are infinite equivalent parameterisations (or rotations) that deliver a minimally restrictive configural model. This is the well-known issue of factor indeterminacy, which arises due to the lack of natural units of measurement for the latent factors being assessed.

@@

\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$ for two groups $A$ and $B$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\ei
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and thresholds to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$). The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}. Still, all of these parameterisations are statistically equivalent. The measurement invariance analysis in this paper is based on the Theta parameterisation (WE{\textTheta}), but results are independent on this choice. The restrictions in \eqref{eq:propens}, \eqref{eq:thresh}, and \eqref{eq:thetapar} define the so-called \emph{configural} model.

@@

Any comparison between socioemotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts. In other words, the items in our new scale must measure externalising and internalising socioemotional skills in the same way in the BCS and MCS data. This property is denominated measurement invariance \citep{Vandenberg2000a,Putnick2016}.

In the framework of factor analysis, measurement invariance is a formally testable property. In this paper, we follow the recent identification methodology by \citet{Wu2016a}. The configural model defined above serves as the starting point. Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models. These models place increasing restrictions on the item parameters, constraining them to be equal across groups. Their fit is then compared to that of the configural model. Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved.

In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).

@@

Let's consider examples from our application. A \emph{loading and threshold invariance} model restricts every item's loading $\lambda$ and threshold $\tau$ parameters to have the same value in the two cohorts. It assumes that the items in our scale have the same relationship with latent skills across the two cohorts. In other words, items have the same salience, or informational content relative to skills. If this model fits as well as the configural model, we can be confident that the socioemotional skills of children in the two cohorts can be placed on the same scale, and their \emph{variances} can be compared. To see why, consider equation \eqref{eq:propens}. If the loading matrix $\bm{\Lambda}$ is the same across cohorts, any difference in latent skills $\Delta\bm{\theta}$ will correspond to the same difference in latent propensities $\Delta\bm{X}^*$. Equality of thresholds $\tau$ ensures that propensities $\bm{X}^*$ map into observed items $\bm{X}$ in the same way.

A \emph{loading, threshold, and intercept invariance} model additionally restricts every item's intercept $\nu$ across cohorts. A good relative fit of this model indicates that socioemotional skills can be compared across cohorts in terms of their \emph{means} as well. To see why, consider the following. Since the $\lambda$ and $\nu$ parameters are the same across cohorts, a child in the BCS cohort with a given level of latent skills $\bar{\bm{\theta}}$ will have the same expected latent item propensities $\bm{X}^*$ as a child with the same skills in the MCS cohort. Again, equality of thresholds $\tau$ fixes the mapping between $\bm{X}^*$ and $\bm{X}$.\footnote{We recognise that simultaneous invariance of \emph{all} items is not the minimum requirement for comparability. In theory, the availability of just one invariant item (known as `anchor') would suffice to fix the scale and location of the system. However, partial invariance approaches are hard to implement in practice. Its validity hinges on selecting one (or more) truly invariant anchor, which is challenging on an a priori basis. The full procedure, restricting all parameters of a certain type across groups, does not identify which items are at the source of the invariance. Algorithms have been proposed to deal with this issue \citep{Yoon2007,Cheung2012}, however there are still doubts on their robustness and their applicability to the categorical case \citep{Vandenberg2016}.}

@@

Again, just like in the broader context of measurement invariance, most evidence regarding the performance of AFIs pertains to scenarios with continuous measures. The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice. Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989}. Subsequent simulation studies -- e.g. \citet{Chen2007} and \citet{Meade2008} -- have updated these thresholds for the continuous case. In particular, \citet{Chen2007} shows in two Monte Carlo studies that the standardised root mean square residual (RMSR) is more sensitive to lack of invariance in factor loadings than in intercepts or residual variances, while the CFI and RMSEA are equally sensitive to all three types of lack of invariance; he suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta SRMR $> .010$.

However, it is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citep{Lubke2004}. Recent studies have advanced simulation evidence around the performance of AFIs in measurement invariance analysis with categorical measures. \citet{Sass2014} finds that the cutoffs from \citet{Chen2007} might not generalise well to problems estimated by WLSMV, but this is mostly confined to smaller sample sizes and detection of small degrees of non-invariance. More recently, \citet{Rutkowski2017} find that a \textDelta RMSE threshold of $.010$ is appropriate for testing equality of slopes and thresholds when the sample size is large, like in our case.

In any case, we present a range of fit indices to provide a more holistic assessment of measurement invariance. The fit of each nested model is compared in Panel A of \autoref{tab:fit}. The overall fit of the configural model is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$. As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels. The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs. In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance. Invariance of loadings and thresholds across cohorts implies that items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.

@@

To summarise, our measurement invariance analysis shows partial comparability of socioemotional skills across cohorts. In particular, the variance of skills can be compared across cohorts, but mean cohort differences do not necessarily reflect differences in skills. We can use scores from our scale to compare children within the same cohort-gender group, but not across cohorts. However, we can also compare within-cohort differences between groups of children, across cohorts. As an example, consider two groups of children A and B in the BCS cohort, and two groups of children C and D in the MCS. We cannot compare the mean level of skills between groups A and C, but we can compare the mean difference between groups A and B with the mean difference between groups C and D. This is the approach we take for the rest of the paper. Refraining from direct cross-cohort comparisons, we interpreting significance and magnitude of within-cohort differences across the cohorts.

>>>>

We use information from two nationally representative longitudinal studies in the UK, which follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) and the Millennium Cohort Study (MCS). The BCS includes all individuals born in Great Britain in a single week in 1970. The cohort members' families -- and subsequently the members themselves -- were surveyed on multiple occasions. For this paper we augment the information collected at the five-year survey with data from birth, adolescence (16), and adulthood (30, 38, 42). The MCS follows individuals born in the UK between September 2000 and January 2002. We use the first survey -- carried out at 9 months of age -- and the sweeps at around 5 and 14 years of age.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}

@@

Ideally, we would compare socio-emotional skills alongside cognitive skills. However, the cognitive tests administered to each cohort have no overlap, even at the item level. We thus use the available cognitive tests in each cohort to estimate simple confirmatory factor models with a single latent dimension, separately by cohort (see \autoref{tab:harmvar} for the tests used). Unlike the other indicators in our analysis, cognitive skills are thus not comparable across cohorts.

Another complication arises from the fact that, differently from the British Cohort Study, the Millennium Cohort Study has a stratified design. It oversamples children living in administrative areas characterised by higher socioeconomic deprivation and larger ethnic minority population \citep{Plewis2007}. We rebalance the MCS sample to make it nationally representative by excluding from the analysis a fraction of observations from the oversampled areas, proportionally to their sampling probability.\footnote{See Table 5.5 in \citet{Plewis2007}. This choice is mainly driven by software limitations. The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.} Finally, we also restrict our sample to individuals born in England and to cases where there is complete information on socio-emotional skills at five years of age. The final sample contains 9,545 individuals from the British Cohort Study, and 5,572 from the Millennium Cohort Study. Summary statistics for the full and estimation samples are displayed in \autoref{tab:sumstats}. After the rebalancing step, the MCS estimation sample closely mirrors the full sample in terms of average observable characteristics, thus preserving representativeness.

@@

Child socio-emotional skills are an unobservable and difficult to measure construct. Over recent years, the measurement of such skills has evolved and, over time, different measures have been used.  As we discuss below, this makes the comparison  of socio-emotional skills across different groups, assessed with different tools, difficult.

A common approach to infer a child's socio-emotional development is based on behavioural screening scales. As part of these tools, mothers (or teachers) indicate whether their children exhibit a series of behaviours -- the \emph{items} of the scale. In the British and Millennium Cohort Studies, two different scales were employed. In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an update to the Rutter scale. It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}. \citet{Goodman1997} administered both scales to a sample of children, and showed that the scores are highly correlated, and the two measures do not differ in their discriminatory ability. The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively. In the child psychiatry and psychology literatures, the Rutter and SDQ scales are regarded as measures of behavioural problems and mental health. However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{Goodman2011}.

While the Rutter and SDQ scales are similar in their components (since the latter was developed from the former, see \citet{Goodman1994}), there is no a priori reason to expect them to be directly comparable. First, the overlap of behaviours described in the two scales is only partial, given that - by design - the SDQ includes also strengths, in addition to weaknesses. Second, the wording of each item is slightly different, both in the description and in the options that can be selected as answers. Third, the different ordering of the items within each scale might lead to order effects. Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2006) might differ due to a host of evolving societal norms. Nonetheless, the level of comparibility of the two scales is higher than that of other scales used in comparative work in the literature reviewed in section \ref{sec:lit}.

As our goal is to compare socio-emotional skills across the two cohorts, we construct a new scale by retaining the items that are worded in a similar way across the two original Rutter and SDQ scales, and making some slight coding adjustments to maximise comparability. In what follows, we will consider the included items to be the same \emph{measure} in the two cohorts. The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability. We exclude from the analysis items that were completely different between the two questionnaires to maximise comparability between the two cohorts, as it is standard good practice in the psychometric literature (see for example \citet{Kern2014}).\footnote{Of course, we could have included them in the factor analysis and treated them as missing in the cohort where they were not administered.} More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.

Item-level prevalence by cohort and gender is in \autoref{tab:meantab}. We see that, in general, there are more similarities across genders within the same cohort, than across cohorts. For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2006 than in 1975. Regardless, a simple cross-cohort comparison of item-level prevalence is misleading because of changing perceptions and norms about what constitutes problematic behaviour in children. The analysis in section \ref{sec:measinv} tackles this issue.

In the remainder of this section, we analyse the properties of the new scale. Following a common approach, we proceed in two steps. First, we carry out an \emph{exploratory} step, where we study the factor structure of our scale. The aim of this step is to examine the correlation between observed measures in a data-driven way, imposing the least possible assumptions. Here, we establish how many latent dimensions of socio-emotional skills the scale is capturing, and which items of our scale are measuring which dimension. As a second step, we set up a \emph{confirmatory} factor model. This model fixes the number of latent dimensions, and imposes a dedicated measurement structure, based on the insights obtained in the exploratory step. This is the model to which we apply the measurement invariance analysis of \autoref{sec:measinv}.

@@

The original Rutter scale, used in the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}. This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \citet{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}. However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used in the MCS cohort, was instead conceived to have five subscales of five items each. The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}. This five-factor structure has been validated in many contexts \citep{Stone2010}; lower-dimensional structures have been also suggested \citep{Dickey2004}. Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}. Indeed, the internalising and externalising dimensions were introduced in psychology by \citet{Achenbach1966}, who showed that they are the two main factors underlying a wide range of psychological measures; as noted in \citet{Achenbach2016}, more than 75,000 articles have been published on internalizing and externalizing problems.

@@

Measurement invariance analysis necessarily starts from a minimally restrictive model, denominated \emph{configural} model. This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts. The restrictions implied in \eqref{eq:propens} and \eqref{eq:thresh} are not sufficient to identify the parameters of the model: even with these assumptions, there are infinite equivalent parameterisations (or rotations) that deliver a minimally restrictive configural model. This is the well-known issue of factor indeterminacy, which arises due to the lack of natural units of measurement for the latent factors being assessed.

@@

\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor.
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$ for two groups $A$ and $B$), and an additional threshold in the reference items above.
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$.
  \item Set all intercepts $\bm{\nu}$ to zero.
  \ei
\ei
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and thresholds to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$). The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}. Still, all of these parameterisations are statistically equivalent. The measurement invariance analysis in this paper is based on the Theta parameterisation (WE{\textTheta}), but results are independent on this choice. The restrictions in \eqref{eq:propens}, \eqref{eq:thresh}, and \eqref{eq:thetapar} define the so-called \emph{configural} model.

@@

Any comparison between socio-emotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts. In other words, the items in our new scale must measure externalising and internalising socio-emotional skills in the same way in the BCS and MCS data. This property is denominated measurement invariance (MI) \citep{Vandenberg2000a,Putnick2016}.

In the framework of factor analysis, measurement invariance is a formally testable property. In this paper, we follow the recent identification methodology by \citet{Wu2016a}. The configural model defined above in section \ref{sec:config} serves as the starting point. Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models. These models place increasing restrictions on the item parameters, constraining them to be equal across groups. Their fit is then compared to that of the configural model. Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved.

In the case where the available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).

@@

Let's consider examples from our application. A \emph{loading and threshold invariance} model restricts every item's loading $\lambda$ and threshold $\tau$ parameters to have the same value in the two cohorts. It assumes that the items in our scale have the same relationship with latent skills across the two cohorts. In other words, items have the same salience, or informational content relative to skills. If this model fits as well as the configural model, we can be confident that the socio-emotional skills of children in the two cohorts can be placed on the same scale, and their \emph{variances} can be compared. To see why, consider equation \eqref{eq:propens}. If the loading matrix $\bm{\Lambda}$ is the same across cohorts, any difference in latent skills $\Delta\bm{\theta}$ will correspond to the same difference in latent propensities $\Delta\bm{X}^*$. Equality of thresholds $\tau$ ensures that propensities $\bm{X}^*$ map into observed items $\bm{X}$ in the same way.

A \emph{loading, threshold, and intercept invariance} model additionally restricts every item's intercept $\nu$ across cohorts. A good relative fit of this model indicates that socio-emotional skills can be compared across cohorts in terms of their \emph{means} as well. To see why, consider the following. Since the $\lambda$ and $\nu$ parameters are the same across cohorts, a child in the BCS cohort with a given level of latent skills $\bar{\bm{\theta}}$ will have the same expected latent item propensities $\bm{X}^*$ as a child with the same skills in the MCS cohort. Again, equality of thresholds $\tau$ fixes the mapping between $\bm{X}^*$ and $\bm{X}$.\footnote{We recognise that simultaneous invariance of \emph{all} items is not the minimum requirement for comparability. In theory, the availability of just one invariant item (known as `anchor') would suffice to fix the scale and location of the system. However, partial invariance approaches are hard to implement in practice. Its validity hinges on selecting one (or more) truly invariant anchor, which is challenging on an a priori basis. The full procedure, restricting all parameters of a certain type across groups, does not identify which items are at the source of the invariance. Algorithms have been proposed to deal with this issue \citep{Yoon2007,Cheung2012}, however there are still doubts on their robustness and their applicability to the categorical case \citep{Vandenberg2016}.}

@@

Again, just like in the broader context of measurement invariance, most evidence regarding the performance of AFIs pertains to scenarios with continuous measures. The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice. Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989}. Subsequent simulation studies -- e.g. \citet{Chen2007} and \citet{Meade2008} -- have updated these thresholds for the continuous case. In particular, \citet{Chen2007} shows in two Monte Carlo studies that the standardised root mean square residual (RMSR) is more sensitive to lack of invariance in factor loadings than in intercepts or residual variances, while the CFI and RMSEA are equally sensitive to all three types of lack of invariance; he suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta RMSR $> .010$.

However, it is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citep{Lubke2004}. Recent studies have advanced the simulation-based evidence on the performance of AFIs in measurement invariance analysis with categorical measures. \citet{Sass2014} find that the cutoffs from \citet{Chen2007} might not generalise well to problems estimated by WLSMV, but this is mostly confined to smaller sample sizes and detection of small degrees of non-invariance. More recently, \citet{Rutkowski2017} find that a \textDelta RMSE threshold of $.010$ is appropriate for testing equality of slopes and thresholds when the sample size is large, like in our case.

In any case, we present a range of fit indices to provide a more complete assessment of measurement invariance. The fit of each nested model is compared in Panel A of \autoref{tab:fit}. The overall fit of the configural model is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$. As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels. The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs. In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance. Invariance of loadings and thresholds across cohorts implies that the items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.

\textbf{[ADD NEW RESULTS]}

@@

To summarise, our measurement invariance analysis shows partial comparability of socio-emotional skills across cohorts. In particular, the variance of skills can be compared across cohorts, but mean cohort differences do not necessarily reflect differences in skills. We can use scores from our scale to compare children within the same cohort-gender group, but not across cohorts. However, we can also compare within-cohort differences between groups of children, across cohorts. As an example, consider two groups of children A and B in the BCS cohort, and two groups of children C and D in the MCS. We cannot compare the mean level of skills between groups A and C, but we can compare the mean difference between groups A and B with the mean difference between groups C and D. This is the approach we take for the rest of the paper. Refraining from direct cross-cohort comparisons, we interpreting significance and magnitude of within-cohort differences across the cohorts.
